---
title: Quick Intro to ggplot
author: Justin Millar
date: '2017-02-22'
slug: quick-intro-to-ggplot
categories:
  - Meetup Presentation
tags:
  - ggplot
  - graphing
description: ''
thumbnail: ''
---

R is capable of creating a wide variety of graphics and data visualizations right out of the box. It’s one of the main reasons for using R. However, base R graphics are often not the most aesthetically pleasing, and can be diffuclt to customize (especially if you’re new to programs like R that don’t use graphic interfaces).

Thankfully there are many packages for making elegant data visualizations in R, and a standout among them is ggplot2. This package is among the most downloaded on CRAN, it won’t take long before you see ggplots everywhere.

First lets install and load the package,

```{r, message=FALSE, warning=FALSE}
require('ggplot2')
library(ggplot2)
```


Let's make a basic histogram using base graphics and ggplot. We'll use the diamonds dataset (this is built into the package).


```{r, message=FALSE, warning=FALSE}
hist(diamonds$carat)
ggplot(data = diamonds) +
  geom_histogram(aes(x = carat))
```


These code for these graphics are similar, as are the outputs, however there are some noticeable differences.


